NHL Realignment Effects on Playoff Qualification

### 2020-12-14 Paper.Rmd
Update Paper.Rmd
added 2017-2018 info to introduction and results, did some minor proof reading to try and make it more cohesive.

>>>>

After doing 10,000 simulations of full NHL seasons using the Monte Carlo simulation methods detailed above, the results can be seen in the figures below. The red shaded graph represents the gap between the East and West 8th seeds (conference gap) under the conference and division alignment in the 2013-14 NHL season while the blue graph represents the gap between the East and West 8th seeds (conference gap) under the conference and division alignment in the 2012-2013 NHL season. The mean conference gap for the new alignment (2013-2014) is 2.30 points while the mean conference gap for the old alignment (2012-2013) is practically zero at -0.02 points.

![](Images/GapVis.png){width=85%}

When looking at the 2017-2018 season we can see something close to what we hypothesized in our introduction, our conference gap does indeed lie somewhat close to halfway between the estimate for the 2013-2014 season and 2012-2013. When rerunning the analysis for 2017-2018 our estimate for conference gap under the 2012-2013 structure was 0.038 which is close to what was estimated above and makes sense as we expect the conferences to be equal in terms of likliehood of making the playoffs under these rules. The mean conference gap across our simulations for the 2017-2018 season was 1.091 which is approximately halfway between 0 and 2.3. It should be noticed that visually it can be seen that the density for the 2017-2018 season conference gap more closesly resembles the 2012-2013 density than the 2013-2014 density did.

![](Images/gapvis20172018.png){width=85%}

Looking at these results, the numbers fall in line with what would be the logical expectation under these circumstances. The 2013-2014 structure moves the balance of teams over towards the East with 16 teams while the West sits at 14 teams. With more teams in the East, it makes sense that the East would have more teams with higher number of total points from the season. To look at these results in a different light, it is key to translate these numbers to the point scoring system of the NHL. On average, the difference between the conference gaps of the two division alignments (2012-2013,2013-2014) comes out to be 2.32 in favor of the 2013-2014 alignment/structure. Under the 2017-2018 analyis our difference in conference gap between alignments comes out to be 1.053 (1.091-0.038). This result makes sense since the conferences are still unbalanced with the East having more teams than the West (16 and 15 respectively) but less unbalanced than 2013-2014 (16 and 14 respectively)  The conference gap has been calculated as:

East 8th Seed Season Total Points - West 8th Seed Season Total Points

The mean conference gap, in terms of total points, means that the East’s 8th seeded team has on average 2.30 more points than the West’s 8th seeded team for 2013-2014, similarly we expect the 8th seed in the Eastern conference to have 1.091 more points than their Western conference counterpart under the 2017-2018 structure. In the NHL, a team earns 2 points for a win, 1 point for an overtime loss, and 0 points for a loss. With this point system,under the 2013-2014 alignment the East’s 8th seeded team on average needs to win one more game during the season or exchange two losses for two losses in overtime. This favors Western conference teams in the way they have to earn less total points throughout the season to make the playoffs than Eastern conference teams. This same bias towards Western conference teams to make the playoffs is present under the 2017-2018 bias but the practical effects of a one point difference is not even worth a single win in an 82 game season, so although this effect is present it's not as drastic.

Not only were the 8th seeds from each conference calculated but each seed for the whole NHL. These box and whisker plots show a wide range of information for the total points for each seed across the 10,000 simulated seasons for both the 2013-2014 and 2017-2018 alignemnts. The endpoints of the whiskers represent the outlier point totals earned by each seed. The sides of the box represent the 2.5% and 97.5% quantiles. The line within the box represents the mean. The structure of this plot makes sense with the curve of less points the lower seed and more points the higher seed the team achieves. The higher the number on the x axis the lower seed of the team; 30 being the worst team in the league (31 for 2017-2018 simulation) and 1 being the best team in the league. Interesting feautures of these plots are a tight spread of simulated points for teams seeded around the middle of the league. This makes sense given that these point values are probably closer to the mean point total for a season and thus more likely under the model we defined. Another note is that the overall range of point totals is greater in the 2013-2014 simulation than it is under the 2017-2018 simulation. This has been compared with real NHL data and the shape lines up in a similar fashion.

## 2013-2014 Points by Seed
![](Images/SeedPlace.png){width=80%}

## 2017-2018 Points by Seed
![](Images/seedplace20172018.png){width=80%}

# Conclusion

Since the change to the conference structure of the NHL to be unbalanced, there have been questions on whether these structures are giving an unfair advantage for one of the conferences: more specifically the Western Conference. With the West having two less teams under the 2013-2014 alignment, it was inferred that each of those teams would have a better chance of making the playoffs. They wouldn’t need to earn as many points to secure a playoff spot as teams in the Eastern Conference. After running 10,000 Monte Carlo simulations of the 2013-2014 NHL season, the results support the case that the Western Conference has a noticeable advantage over the Eastern Conference. The 8th seed in the east on average has an end of season point total of 2.3 (1.091 for 2017-2018) points higher than the 8th seed in the west. This can cause problems because a statistically worse team can make the playoffs in the Western Conference while a better team might miss out in the Eastern Conference. This is support of the idea that these rules introduce a bias agaisnt teams in the Eastern conference which goes agaisnt the league being "fair" to all teams.
